# Summary statistics

## Mathmatics

### Mean

The mean is generally presented in statistics as $\overline{X}$. The formula for the mean is:

$\overline{X} = \frac{\sum_i^n{x_i}}{n}$

Essentially, all data points are added together and then divided by the total number of data points.

For example, say you had a dataset of 2, 3, 4, and 7

you would add these together which would equal 16.

Then you would divide 16 by the number of data points (4).
So, the mean of this dataset would be 4


### Median

*to be completed


### Variance

*to be completed


### Standard deviation

*to be completed


# In R

You can work out the mean of a dataset in R in a variety of ways.

```{r}
mean_dataset <- c(8,9,11,13,14) # lets say we have a dataset

mean_dataset # lets view the data points

8 + 9 + 11 + 13 + 14 # we now know it is 55

55 / 5 # and we now know the mean is 11


```
The above example is fine if you have a few numbers, but what if your dataset is larger?
Well, you can use certain packages to make this easier

```{r , warning=FALSE, message=FALSE}
library(tidyverse) # lets load our library
```

```{r}
head(mpg) # we are using the mpg dataset that is built into the tidyverse package
```
lets say we want to work out the mean year that cars are manufactured

```{r}
mpg %>% # pipe function
  summarise(mean_year = mean(year)) # summarise function in dplyr package
```
Now we know that the approximate mean year that cars are manufactured is around 2003-04.


